---
title: "Create ACLED country code vector"
author: "Spencer Graves"
date: "2024-12-09"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::knitr}
  %\VignetteIndexEntry{ACLEDctryCode}
  %\usepackage[UTF-8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This vignette merges the `ctry` and `WBdat` datasets 
created by the `readACLEDsummaryData` and 
`readWorldBankData` vignettes and creates a named 
character vector `ACLEDctryCode`  to facilitate 
translation between the "country" or geographic 
entity included in `ACLED` data and the corresponding 
3-letter [ISO 3166 country code](https://en.wikipedia.org/wiki/List_of_ISO_3166_country_codes).

We proceed as follows: 

1. Access `data(ctry)` and `data(WBdat)`. 
2. Create `ACLEDctryCode empty as character(nrow(ctry))`. 
3. Fill `ACLEDctryCode` with the `WBdat CtryCode` for every `WBdat Country` in the ACLED data `ctry`. 
4. Find the desired 3-letter ISO codes in [`rworldmap::countrySynonyms`](https://search.r-project.org/CRAN/refmans/rworldmap/html/countrySynonyms.html) for for countries NOT in `WBdat` with the name used by ACLED. 
5. Manually find the desired 3-letter codes for other ACLED countries in `ctry` that steps 3 and 4 did not find. 
6. Save. 

## 1. `data(ctry)` and `data(WBdat)`

```{r load_ctry}
library(acledr)
data(ctry)
data(WBdat)
##
```

## 2. Create `ACLEDctryCode`

```{r createACLEDcc}
ACLEDctryCode <- character(nrow(ctry))
names(ACLEDctryCode) <- ctry$Country
```

## 3. Fill `ACLEDctryCode` from `WBdat`

```{r ctryCdFrWB}
str(ACLEDinWB <- grepInTable(ctry$Country, 
                              WBdat[4:3]))
ACLEDctryCode[names(ACLEDinWB)] <- ACLEDinWB

NnchACLEDinWB <- table(nchar(ACLEDctryCode))
nNotInWB <- (sum(NnchACLEDinWB) - NnchACLEDinWB[1])

str(ACLEDnotInWB <- which(nchar(ACLEDctryCode)!=3))
```

## 4. Find ISO3 in `countrySynonyms` for ACLED countries not in `WBdat`

```{r notInWB}
ISO3notInWB <- grepInTable(names(ACLEDnotInWB))
ISO3notWB3 <- (nchar(ISO3notInWB)==3)
ACLEDctryCode[ACLEDnotInWB[ISO3notWB3]] <- 
                      ISO3notInWB[ISO3notWB3]
(mysteryCtries <- names(ACLEDnotInWB[!ISO3notWB3]))
table(nchar(ACLEDctryCode))
```

## 5. Manually find `ISO3` for ACLEd countries not yet found 

The Wikipedia article on ["List of ISO 3166 country codes"
](https://en.wikipedia.org/wiki/List_of_ISO_3166_country_codes) says, "Caribbean Netherlands – See Bonaire, Sint Eustatius and Saba", for which `ISO3 = BES`. It also says that [Christmas Island](https://en.wikipedia.org/wiki/Christmas_Island) is a territory of Australia with `ISO3 = CXR`. 

Let's add those codes to `ACLEDctryCodes`.

```{r ACLEDctryCode9}
ACLEDctryCode['Caribbean Netherlands'] <- 'BES'
ACLEDctryCode['Christmas Island'] <- 'CXR'
str(ACLEDctryCode)
head(ACLEDctryCode)
(NnchACLEDctCd <- table(nchACLEDctCd <- nchar(ACLEDctryCode)))
(ACLEDctNotFound <- sum(NnchACLEDctCd[-1]))
```

Done: We now have `ISO3` codes for all `ACLEDctries1`. 

## 6. Save 

Save `ACLEDpopGDP`. Execute ONLY if you actually want to update the data in the package (or are running this vignette in a context outside of a pacakge like `acledr`).  

```{r save}
# usethis::use_data(ACLEDctryCode, overwrite=TRUE)
```

