---
title: "Read ACLED summary data"
author: "Spencer Graves"
date: "2024-12-06"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::knitr}
  %\VignetteIndexEntry{Read ACLED summary data}
  %\usepackage[UTF-8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

[Armed Conflict Location and Event Data (`ACLED`)](https://en.wikipedia.org/wiki/Armed_Conflict_Location_and_Event_Data) is a non-profit organization specializing in collecting incident-level data on conflicts that support analysis and crisis mapping in near real time. ACLED codes the dates, actors, locations, fatalities, and types of all incidents of political violence and demonstration events around the world as reports appear in the literature they monitor. The project began in 2005 focusing on African states. Its coverage expanded multiple times until 2021-01 when it covered the entirety of humanity, by which time it had over 1.3 million events in its database. [A pdf file gives the start data of coverage for different countries.](https://acleddata.com/acleddatanew/wp-content/uploads/dlm_uploads/2019/01/ACLED_Country-and-Time-Period-coverage_updatedFeb2022.pdf) These range from 1997-01 for continental Africa to 2021-01 for most of North America and Oceana (including Australia and New Zealand) and Antarctica. 

The [ACLED Explorer](https://acleddata.com/explorer/) makes it moderately easy for someone to download counts of either events or fatalities in a specific date range by "political violence", "organized violence", "demonstrations", "civilian targeting" or "all event types". Data can be grouped by "region", "country", `Admin1` and "Location". Counts can optionally be restricted to only specific actors or "actor types". Events can optionally be grouped by event type and by day or week. 

Long term, we'd like to build models of the evolution of the numbers of events and deaths by event type, month and country. Such models have two uses: 

- First to [compute the probability of detecting a reduction in the lethality of a particular conflict in a specific period of time if an intervention had an effect of a given magnitude (statistical power)](https://en.wikipedia.org/wiki/Power_(statistics)). 
- Then if a proposed intervention is actually implemented, the same models can be used to evaluate the efficacy of the intervention(s).

Before we do that, we first attempt simpler tasks: 

1. Read and summarize `ACLED` data in various ways. *[this vignette]*
2. Read World Bank (`WB`) data on population and level of econmic activity as Gross Domestic Product per capita (`GDPpc`) at both Purchasing Power Parity (`PPP`) and nominal: We need `PPP` to evaluate the impact of `GDPpc` and changes in `GDPpc` on the level of conflict. And we need nominal `GDP` to estimate budgets for certain types of interventions. (`GDP` is, of course, population times `GDPpc`.) 
3. Merge `ACLED` and `WB` data. 
4. Build static models of the relationships between `ACLED` and `WB` variables. 

## Extract, download and read `ACLED` data 

To make everything comparable, we downloaded events and fatalities from ACLED Explorer by country and event type for each month starting with 2021-01 as `*.csv` files. This involved two requests from "ACLED Explorer" for "Event Counts" separate from "Fatality Counts". For each extraction, we requested, (2) "All event types", (3) a custom date range starting with the first day of the first month requested and ending with the last day of the last month requested. We also requested (4) "Select All" regions and (5) countries. (6) We did NOT request "specific Admin1 units". (7) We asked to group the data by country. (8) We did NOT select specific actors or types of actors. But we DID want to see events groups by (9) event type and (10) time period: month. 

We moved these two files in the current working directory and renamed them `acled_aggregated_2021events.csv` and `acled_aggregated_2021fat.csv`, respectively. 

These files were 2.5 MB each. For CRAN, ["As a general rule, neither data nor documentation should exceed 5MB"](https://www.thecoatlessprofessor.com/programming/r/size-and-limitations-of-packages-on-cran/). To meet these limits, sample data from 2023-12-01 to 2024-01-31 were extracted to include in the `extdata` directory of the package. 

```{r acledDir}
#extdata <- pkgload::shim_system.file('extdata', package='acledr')
# Error: 'shim_system.file' is not an exported object from 
#        'namespace:pkgload'
if(require(acledr)){ 
  acledDir <- fs::path_package('acledr')
} else acledDir <- ''

if((length(acledDir)<1) || (acledDir == '')){ 
  wd <- getwd()
  wdList <- strsplit(wd, '/')[[1]]
  acledri <- which(wdList=='acledr')
  acledDir <-  do.call(file.path, as.list(wdList[1:acledri]))
}

acledDir
```

Got `acledDir`. This is difficult for `*.Rmd` in the vignettes folder. 

```{r extdata}
inst <- dir(acledDir, pattern='^inst$', full.names=TRUE)
if((length(inst)<1) || (inst=='')){
    extdata <- dir(acledDir, pattern='^extdata$', 
                 full.names=TRUE)
} else {
    extdata <- dir(inst, pattern='^extdata$', 
                 full.names=TRUE)
}
extdata
```

Now find the deaths and events files. 

```{r d_e_files}
csvFiles <- dir(extdata, pattern='\\.csv$')
(deathFile <- dir(extdata, pattern='fat\\.csv$', 
                  full.names = TRUE))
(eventFile <- dir(extdata, pattern='events\\.csv$', 
                  full.names = TRUE))
```

Let's read these two files and confirm that they are identical except for the last column. 

```{r readDat}
deaths <- read.csv(deathFile)
events <- read.csv(eventFile)

all.equal(deaths[-4], events[-4])
```

Looks good. Let's combine them to create `ctryMoEv`. We then summarize those 

- over `Event.month` to create `ctryEv` 
- and across `Event.Type` to create `ctryMo`. 

Then summarize `ctryEv` further 

- across `Country` to get `Ev` 
- and across `Event.Type` to get `ctry`. 

```{r ctryMoEv}
ctryMoEv <- events
ctryMoEv$Event.month <- as.Date(events$Event.month)
(firstMo <- min(ctryMoEv$Event.month))
(lastMo <- max(ctryMoEv$Event.month))

ctryMoEv$Deaths <- deaths$Fatalities
ctryMoEv$deathsPerEvent <- (deaths$Fatalities/events$Events)
str(ctryMoEv)
```

Before summarizing, let's compute `nCtries`, `nMonths`, and `nTypes`: 

```{r factorNobs}
Ctries <- table(ctryMoEv$Country)
(nCtries <- length(Ctries))

Months <- table(ctryMoEv$Event.month)
(nMonths <- length(Months))

Types <- table(ctryMoEv$Event.Type)
(nTypes <- length(Types))

(Nobs <- nrow(ctryMoEv))
all.equal(Nobs, nCtries*nMonths*nTypes )
```

### `ctryEv` <- Sum (`Events`, `Deaths`) by (`Country`, `Event.Type`) across `Event.month`

```{r ctryEv}
str(Nmo <- table(ctryMoEv$Event.month))
table(Nmo)

library(dplyr)
ctryEv <- (ctryMoEv %>% group_by(Country, Event.Type) 
           %>% summarize(Events=sum(Events), Deaths=sum(Deaths)))
ctryEv$DeathsPerEvent <- with(ctryEv, Deaths/Events)
dim(ctryEv)
head(ctryEv, 2)
tail(ctryEv, 2)
```

### `ctryMo` <- sum (`Events`, `Deaths`) by (`Country`, `Event.month`) across `Event.Type`

```{r ctryMo}
ctryMo <- (ctryMoEv %>% group_by(Country, Event.month) 
           %>% summarize(Events=sum(Events), Deaths=sum(Deaths)))
ctryMo$DeathsPerEvent <- with(ctryMo, Deaths/Events)
dim(ctryMo)
head(ctryMo, 2)
tail(ctryMo, 2)
```

### `Ev` <- sum (`Events`, `Deaths`) by `Event.Type` 

```{r Ev}
Ev <- (ctryEv %>% group_by(Event.Type) 
           %>% summarize(Events=sum(Events), Deaths=sum(Deaths)))
Ev$DeathsPerEvent <- with(Ev, Deaths/Events)
dim(Ev)
Ev
```

### `ctry` <- sum (`Events`, `Deaths`) by `Country` 

```{r ctry}
ctry <- (ctryEv %>% group_by(Country) 
           %>% summarize(Events=sum(Events), Deaths=sum(Deaths)))
ctry$DeathsPerEvent <- with(ctry, Deaths/Events)
ctry <- as.data.frame(ctry)
rownames(ctry) <- ctry$Country
dim(ctry)
head(ctry, 2)
tail(ctry, 2)
```

Looks sensible. 

## Save 

Save `ACLEDpopGDP`.

```{r}
#| label: save
#save(ACLEDpopGDP, file='ACLEDpopGDP.rda', 
#     compress=TRUE)
```

